Towards a Wide-Coverage Grammar for Swedish Using GF

### 2012-01-06 doc/report/report.tex
report

Parsing, that is the task of automatically identifying morphological and
syntactical structures, is subject to increasing interest, especially
considering the steadily growing amount of data available online. 
So far there is no freely available grammar-driven parser that gives a deep
analysis for Swedish. The fact that %the existing rule-based Swedish
a parser or grammar is not freely available does not only restrict its
usage but also its possibilities of being further developed and our
goal is
to create an open source Swedish grammar. % from which we derive a parser
%accepting all sentences described by the given rules.
As a freely available resource, it can be continuously enhanced and
made use of in other projects.\\
To build the grammar from scratch would be not only time consuming but
would also mean that already existing resources would have to be reimplemented.
In order to not reinvent the wheel we proceed from a combination of well-tested
resources.
%TODO verify this
We start from a GF resource grammar, consisting of an abstract and a concrete syntax
file defining a set of rules describing morpology and syntax.
From now on, this is what we mean by grammar, as
opposed to grammar in the traditional linguistic sense. 
From GF, we get %and thereby get
%The usage of GF allows us to start from
a well-defined system for describing the language, as well as a strong connection to
more than 20 other languages implemented in the framework. Further, we use
the extensive lexicon Saldo and the treebank Talbanken.
%>>The grammar defines a parser, which accepts all constructions defined by the grammar.

@@

%> dependent and implicit types : agda. (read gf book and krasses book)
A grammar written in GF can be used for both parsing and generation.
The parsing algorithm is incremental and has polynomial time and space
complexity \cite{gfMech}. 
The GF package also provides various tools for working with and using the grammars:
a compiler, an interpreter and a runtime system.
The grammars can be compiled into portable grammar format (PGF) \cite{pgf},
supported by %There are libraries for using PGF files with
Java, Java script, C and Haskell libraries. 
%>> Haskell only complete, also or C\# and C.  %may also be used in Android applications.
The interpreter, the GF shell, allows
the user to test grammars by commands for parsing, visualization of parse trees,
random generation, word alignment, morphological quizzes etc.
An interactive GF editor and the GF shell can be tried out online 
\footnote{http://www.grammaticalframework.org/demos/gfse/}. See figure \ref{fig:shell},
 The first command will parse the given
sentence and visualize the resulting parse tree. The second will visualize the 
abstract syntax tree. The trees will be showed in the pgf-format by the program evince.
\begin{figure}[h]
\begin{verbatim} 
> parse "jag ser katten" | visualize_parse -format=pdf -view=evince
> parse "jag ser katten" | visualize_tree  -format=pdf -view=evince
\end{verbatim}
\caption{Example of how the GF shell can be used.}\label{fig:shellvp}
\label{fig:shell}
\end{figure}

@@

\section{Results}
\label{sec:saldoRes}
The resulting dictionary contains more than 100 000 entries, approximately 80 \% 
of the total size of Saldo.
There are a number of reasons why some words were not imported,
the most obvious one is that we do not want all categories from
Saldo in the GF lexicon. Prepositions, %(\emph{pp})
numerals %(\emph{nl}),
personal pronouns etc.
are assumed to be present in the resource grammars and should not be added again.
Saldo contains many pronouns which
are not analysed the same way in GF (see section \ref{sec:swegf}).
Befor adding them to our lexicon, we need to do more analysing to find their correct GF-category.
%TODO read again
%They can thus not simply be extracted -- we first need to find their correct GF category.
Some experiments on finding the category
%the right GF-category for those
have been done using Talbanken, see p. \pageref{sec:gf.quant}.

@@

Apart from these restrictions, noun phrases containing reflexive pronouns
may be used as any other \verb-NP-. They may be conjoined 
(\ref{ex:ReflConj} a,b) and used with other determiners (\ref{ex:ReflPredet}).
In the standard GF analysis, which is preformed bottom-up starting from the pos-tags, 
all information about semantical roles are lost. There is no way
of knowing whether a \verb-NP- acts as subject or object. For this reason,
the formalization of reflexive pronouns required the use of a different analysis.
%and used as genitives (sentence \ref{ex:ReflGen}) as such.
%Could be combined with 'sig', a ReflNP without another noun.
In short, what is wanted can be summarized as follows:\\
%\begin{figure}[h]
\begin{enumerate}
\item{
Construct a noun phrase from a common noun phrase \\
katt $\rightarrow$ sin katt} % \\
\item{Modify it like a \verb|NP|\\
sina katter $\rightarrow$ alla sina katter}
\item{Use it only as object\\
sin katt $\rightarrow$ han såg sin katt }
\end{enumerate}
%\caption{Rules for noun phrases with reflexive possessive pronouns}
%\label{fig:reflrestrict}
%\end{figure}

@@

We hence combine the-part-of-speech driven analyse normally preformed by
GF with a part-of-sentence analysis, where the dependent types gives the semantical
information we were missing.

@@

The grammar has been extended and enhanced, and its current status is
a specialized extension of the resource grammar.
Besides parsing, the grammar may well be used for language generation,
which is fast even with an extensive lexicon.
%TODO make nice!
Although it cannot not been verified, we believe that the majority of the
sentences generated are grammatically correct in a syntactical point of view.
Without any semantical knowledge, nonsense phrases cannot be avoided in
random generation. However, given that the abstract tree has a meaningful
interpretation,
%TODO move to Misc?
the linearization should be correct. There are some cases when the correctness
of the output has been put aside in order to increase the expressivity, such
as \verb-ComplBareVV-, which use a verb-complement verb without the inifite mark, 
as in sentence (\ref{sent:complVV1}b).
\eenumsentence{\item\shortex{5}
{Jag& börjar&\textbf{att}& bli& hungrig}
{\emph{I}& \emph{begin}& \textbf{\emph{to}}&\emph{become}& \emph{hungry}}
{I'm getting hungry}
\item
\shortex{4}
{Jag& börjar& bli& hungrig}
{\emph{I}& \emph{begin}& \emph{become}& \emph{hungry}}
{I'm getting hungry}}\label{sent:complVV1}
Since there is no information about which verbs that allows the inifinite mark
to be left out, it will also allow the more questionable sentence (\ref{sent:complVV2}b).
\eenumsentence{
\item \shortex{5}
{Jag & vill & \textbf{att}& du & går} 
{I & want & \textbf{to} &you & leave}
{I want you to leave}
\item \shortexnt{4}
{Jag & vill & du & går} 
{I & want &you & leave}}\label{sent:complVV2}
%ComplBareVV, SupCl
For random generation, the sentences generated may be next to impossible
to interpret without looking at the abstract tree, since they may contain
too many attributes, relative clauses etc. 
%\enumsentence{Kommer övriga datorer idag som ska vara inte att ha målats?}
%\label{sent:what}
%Sentence (\ref{sent:what})
%PredVP (DetCN Subject (DetSub oovriga_Det) (RelCN (AdvCN (UseN computer_N) (AdvSub today_Adv)) (UseRCl (TTAnt TFut ASimul) PPos (RelVP IdRP UseCopula)))) (PassVP (SlashV2A paint_V2A (ComparA Object red_A (UsePN Object johan_PN))))
%The use of the NONEXIST pattern, as described in \ref{sec:Formal}. When using
%the grammar for generation, an extra

@@

\begin{itemize}
\item
\textit{Pronominal object shift} is common in Swedish and obligatory in Danish and Norwegian.
\enumsentence{\begin{tabular}[t]{@{}*{11}{l@{\ }}}
a. & Jag & ser&  honom& \textbf{inte}&~~ & b.&Jag &ser &\textbf{inte} &honom\\
   & \emph{I} & \emph{see}&\emph{him}&\emph{not}&&& \emph{I} &\emph{see}&\emph{not}&\emph{him}\\
\end{tabular}}
Personal pronouns are allowed to precede negation and some other sentence adverbials --\textit{satsadverbial} --
in main clauses without auxiliary verbs.

@@

Our final goal is to be able to parse unrestricted text, but considering that
no syntactic theory is yet capable
of wholy covering a natural language, we are satisfied with %theconsider our grammar 
%when used without a robust layer 
the grammar implementation of an important fragment of Swedish.
We have therefore focused on constructions that are frequent in Talbanken, 
and following methods for making the parser robust as described in section \ref{sec:future}
we still believe that it is possible to achieve the goal. 
\\

>>>>

Parsing, that is the task of automatically identifying morphological and
syntactical structures, is subject to increasing interest, especially
considering the steadily growing amount of data available online. 
So far there is no freely available grammar-driven parser that gives a deep
analysis for Swedish. The fact that %the existing rule-based Swedish
a parser or grammar is not freely available does not only restrict its
usage but also its possibilities of being further developed and our
goal is
to create an open source Swedish grammar. % from which we derive a parser
%accepting all sentences described by the given rules.
As a freely available resource, it can be continuously enhanced and
made use of in other projects.\\
To build the grammar from scratch would be not only time consuming but
would also mean that already existing resources would have to be reimplemented.
In order to not reinvent the wheel we proceed from a combination of well-tested
resources.
%TODO verify this
We start from a GF resource grammar, consisting of an abstract and a concrete syntax
file defining a set of rules describing morphology and syntax.
From now on, this is what we mean by grammar, as
opposed to grammar in the traditional linguistic sense. 
From GF, we get %and thereby get
%The usage of GF allows us to start from
a well-defined system for describing the language, as well as a strong connection to
more than 20 other languages implemented in the framework. Further, we use
the extensive lexicon Saldo and the treebank Talbanken.
%>>The grammar defines a parser, which accepts all constructions defined by the grammar.

@@

%> dependent and implicit types : agda. (read gf book and krasses book)
A grammar written in GF can be used for both parsing and generation.
The parsing algorithm is incremental and has polynomial time and space
complexity \cite{gfMech}. 
The GF package also provides various tools for working with and using the grammars:
a compiler, an interpreter and a runtime system.
The grammars can be compiled into portable grammar format (PGF) \cite{pgf},
supported by %There are libraries for using PGF files with
Java, Java script, C and Haskell libraries. 
%>> Haskell only complete, also or C\# and C.  %may also be used in Android applications.
The interpreter, the GF shell, allows
the user to test grammars by commands for parsing, visualization of parse trees,
random generation, word alignment, morphological quizzes etc.
An interactive GF editor and the GF shell can be tried out online 
\footnote{http://www.grammaticalframework.org/demos/gfse/}. See figure \ref{fig:shell},
 The first command will parse the given
sentence and visualize the resulting parse tree. The second will visualize the 
abstract syntax tree. The trees will be showed in the pdf-format by the program evince.
\begin{figure}[h]
\begin{verbatim} 
> parse "jag ser katten" | visualize_parse -format=pdf -view=evince
> parse "jag ser katten" | visualize_tree  -format=pdf -view=evince
\end{verbatim}
\caption{Example of how the GF shell can be used.}\label{fig:shellvp}
\label{fig:shell}
\end{figure}

@@

\section{Results}
\label{sec:saldoRes}
The resulting dictionary contains more than 100 000 entries, approximately 80 \% 
of the total size of Saldo.
There are a number of reasons why some words were not imported,
the most obvious one is that we do not want all categories from
Saldo in the GF lexicon. Prepositions, %(\emph{pp})
numerals %(\emph{nl}),
personal pronouns etc.
are assumed to be present in the resource grammars and should not be added again.
Saldo contains many pronouns which
are not analysed the same way in GF (see section \ref{sec:swegf}).
Before adding them to our lexicon, we need to do more analysing to find their correct GF-category.
%TODO read again
%They can thus not simply be extracted -- we first need to find their correct GF category.
Some experiments on finding the category
%the right GF-category for those
have been done using Talbanken, see p. \pageref{sec:gf.quant}.

@@

Apart from these restrictions, noun phrases containing reflexive pronouns
may be used as any other \verb-NP-. They may be conjoined 
(\ref{ex:ReflConj} a,b) and used with other determiners (\ref{ex:ReflPredet}).
In the standard GF analysis, which is preformed bottom-up starting from the pos-tags, 
all information about semantic roles are lost. There is no way
of knowing whether a \verb-NP- acts as subject or object. For this reason,
the formalization of reflexive pronouns required the use of a different analysis.
%and used as genitives (sentence \ref{ex:ReflGen}) as such.
%Could be combined with 'sig', a ReflNP without another noun.
In short, what is wanted can be summarized as follows:\\
%\begin{figure}[h]
\begin{enumerate}
\item{
Construct a noun phrase from a common noun phrase \\
katt $\rightarrow$ sin katt} % \\
\item{Modify it like a \verb|NP|\\
sina katter $\rightarrow$ alla sina katter}
\item{Use it only as object\\
sin katt $\rightarrow$ han såg sin katt }
\end{enumerate}
%\caption{Rules for noun phrases with reflexive possessive pronouns}
%\label{fig:reflrestrict}
%\end{figure}

@@

We hence combine the-part-of-speech driven analyse normally preformed by
GF with a part-of-sentence analysis, where the dependent types gives the semantic
information we were missing.

@@

The grammar has been extended and enhanced, and its current status is
a specialized extension of the resource grammar.
Besides parsing, the grammar may well be used for language generation,
which is fast even with an extensive lexicon.
%TODO make nice!
Although it cannot not been verified, we believe that the majority of the
sentences generated are grammatically correct in a syntactical point of view.
Without any semantic knowledge, nonsense phrases cannot be avoided in
random generation. However, given that the abstract tree has a meaningful
interpretation,
%TODO move to Misc?
the linearization should be correct. There are some cases when the correctness
of the output has been put aside in order to increase the expressivity, such
as \verb-ComplBareVV-, which use a verb-complement verb without the infinitive mark, 
as in sentence (\ref{sent:complVV1}b).
\eenumsentence{\item\shortex{5}
{Jag& börjar&\textbf{att}& bli& hungrig}
{\emph{I}& \emph{begin}& \textbf{\emph{to}}&\emph{become}& \emph{hungry}}
{I'm getting hungry}
\item
\shortex{4}
{Jag& börjar& bli& hungrig}
{\emph{I}& \emph{begin}& \emph{become}& \emph{hungry}}
{I'm getting hungry}}\label{sent:complVV1}
Since there is no information about which verbs that allows the infinitive mark
to be left out, it will also allow the more questionable sentence (\ref{sent:complVV2}b).
\eenumsentence{
\item \shortex{5}
{Jag & vill & \textbf{att}& du & går} 
{I & want & \textbf{to} &you & leave}
{I want you to leave}
\item \shortexnt{4}
{Jag & vill & du & går} 
{I & want &you & leave}}\label{sent:complVV2}
%ComplBareVV, SupCl
For random generation, the sentences generated may be next to impossible
to interpret without looking at the abstract tree, since they may contain
too many attributes, relative clauses etc. 
%\enumsentence{Kommer övriga datorer idag som ska vara inte att ha målats?}
%\label{sent:what}
%Sentence (\ref{sent:what})
%PredVP (DetCN Subject (DetSub oovriga_Det) (RelCN (AdvCN (UseN computer_N) (AdvSub today_Adv)) (UseRCl (TTAnt TFut ASimul) PPos (RelVP IdRP UseCopula)))) (PassVP (SlashV2A paint_V2A (ComparA Object red_A (UsePN Object johan_PN))))
%The use of the NONEXIST pattern, as described in \ref{sec:Formal}. When using
%the grammar for generation, an extra

@@

\begin{itemize}
\item
\textit{Pronominal object shift} is common in Swedish and obligatory in Danish and Norwegian.
\enumsentence{\begin{tabular}[t]{@{}*{11}{l@{\ }}}
a. & Jag & ser&  honom& \textbf{inte}&~~ & b.&Jag &ser &\textbf{inte} &honom\\
   & \emph{I} & \emph{see}&\emph{him}&\emph{not}&&& \emph{I} &\emph{see}&\emph{not}&\emph{him}\\
\end{tabular}}
Personal pronouns are allowed to precede negation and some other sentence adverbials % --\textit{satsadverbial} --
in main clauses without auxiliary verbs.

@@

Our final goal is to be able to parse unrestricted text, but considering that
no syntactic theory is yet capable
of wholly covering a natural language, we are satisfied with %theconsider our grammar 
%when used without a robust layer 
the grammar implementation of an important fragment of Swedish.
We have therefore focused on constructions that are frequent in Talbanken, 
and following methods for making the parser robust as described in section \ref{sec:future}
we still believe that it is possible to achieve the goal. 
\\
